---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---

*Requires dplyr, pander and ggplot2 packages*
```{r requiremens, message=FALSE}
require(dplyr)
require(ggplot2)
require(pander)
```


## Loading and preprocessing the data
1. Show any code that is needed to Load the data (i.e. read.csv())

```{r load}
activity <- read.csv(file = "activity.csv")
```

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r preprocess}
activity$date <- as.Date.character(x = activity$date, format = "%Y-%m-%d")
```


## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

```{r byDay, message=FALSE}

## saves Data frame with NA Values for later filling
activityNA <- subset(x = activity, is.na(activity$steps))

activity <- activity[complete.cases(activity),]

byDay <- activity %>%
        group_by(date) %>%
        summarise(total.steps = sum(steps),
                  total.mean = mean(steps, na.rm = T),
                  total.median = median(steps[steps>0], na.rm = T))
```


1. Make a histogram of the total number of steps taken each day

```{r histPlot}
## using ggplot2
qplot(data = byDay, x = date, y = total.steps, geom = "col", 
      xlab = "Date", ylab = "Total Steps", main = "Total steps taken each Day")
```


2. Calculate and report the mean and median total number of steps taken per day
  
  
  *has been calculated with the dplyr function*
```{r table, message=FALSE, results = "asis"}
pandoc.table(byDay[,c(1,3,4)])
```


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r byTime, message=FALSE}
byTime <- activity %>%
        group_by(interval) %>%
        summarise(total.steps = sum(steps),
                  mean.steps = mean(steps, na.rm = T))

## using R base plotting 
with(data = byTime, plot(x = interval, y = mean.steps, type = 'l', 
                         main = "Average Steps taken Each 5 min interval", 
                         xlab = "Five Minute Interval", 
                         ylab = "Mean Steps"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r maxSteps}
max.steps <- subset(x = byTime, mean.steps == max(byTime$mean.steps))
```

The **`r max.steps$interval`**  five minute interval contains te maximum number of steps with **`r max.steps$total.steps`** steps and mean **`r  max.steps$mean.steps`**.

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset  
*caculated by the activityNA <- subset(x = activity, is.na(activity$steps)) code*

**Total rows with NA values is: `r nrow(activityNA)`**

2. Devise a strategy for filling in all of the missing values in the dataset.

```{r fillNAs}
## fill NA values with the Average Steps taken Each 5 min interval
## merging the NA data with the by interval data mean steps.

filledActivity <- merge(x = byTime[c(3, 1)], y = activityNA[c(2,3)])
names(filledActivity) <- c("interval", "steps", "date")

```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r filledData, message=FALSE}
## binds the Filled Data Set With the original Data
newActivity <- bind_rows(activity, filledActivity)
newActivity <- arrange(.data = newActivity, newActivity$date)

```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r newData, message=FALSE, results = "asis"}
newByDay <- newActivity %>%
        group_by(date) %>%
        summarise(total.steps = sum(steps),
                  total.mean = mean(steps, na.rm = T),
                  total.median = median(steps[steps>0], na.rm = T))

## uses ggplot2
ggplot(data = newByDay, aes(x = date, y = total.steps))+
        geom_bar(stat="identity", fill="steelblue")+
        theme_minimal()+
        labs(title = "Total steps taken each Day")+
        labs(x = "Date", y = "Total Steps" )

pandoc.table(newByDay[,c(1,3,4)])
```
  
Do these values differ from the estimates from the first part of the assignment? **YES**   

What is the impact of imputing missing data on the estimates of the total daily number of steps? Main diference is that it adds data to ``r unique(activityNA$date)`` days

## Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r weekdays}
## adds weekday collumn
newActivity$weekday <- weekdays(abbreviate = T, newByDay$date)
## adds factor
newActivity$period <- factor(
        ifelse(newActivity$weekday %in% c("sáb", "dom"), "weekend", "weekday"))
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r weekdaysPlot}
qplot(x = interval, y = steps, data = newActivity, facets = period~., 
      geom = "line")
```

